---
title: "One-Dimensional Data in R"
author: "PLSC 31101"
date: "Fall 2019"
output: html_document
---

## Vectors 

Let's start with one-dimensional (1d) objects. There are two kinds:

1. **Atomic vectors** - also called, simply, **vectors**. 
2. **Lists**: Lists are distinct from atomic vectors because lists can contain other lists.

We'll discuss **atomic vectors** first. 

### Creating Vectors

Vectors are 1-dimensional chains of values. We call each value an _element_ of a vector. 

Atomic vectors are usually created with `c()`, which is short for 'combine':

```{r}
x <- c(1, 2, 3)
x
length(x)
```

We can also add elements to the end of a vector by passing the original vector into the `c` function, like so:

```{r}
z <- c("Beyonce", "Kelly", "Michelle", "LeToya")
z <- c(z, "Farrah")
z
```

Notice that vectors are always flat, even if you nest `c()`'s:

```{r}
# these are equivalent
c(1, c(2, c(3, 4)))
c(1, 2, 3, 4)
```

### Naming a Vector

We can also attach names to our vector. This helps us understand what each element refers to.

You can give a name to the elements of a vector with the `names()` function. Have a look at this example:

```{r}
days_month <- c(31, 28, 31, 30, 31, 30, 31, 31, 30, 31, 30, 31)
names(days_month) <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

days_month
```

You can name a vector when you create it:

```{r}
some_vector <- c(name = "Rochelle Terman", profession = "Professor Extraordinaire")
some_vector
```

Notice that in the first case, we surrounded each name with quotation marks. But we don't have to do this when creating a named vector. 

Names don't have to be unique, and not all values need to have a name associated. However, names are most useful for subsetting, described in the next chapter. When subsetting, it is most useful when the names are unique. 

### Calculations on Vectors

One of the most powerful things about vectors is that we can perform arithmetic calculations on them.

For example, we can sum up all the values in a numerical vector using __sum__:

```{r}
a <- c(1, -2, 3)
sum(a)
```

We can also sum two vectors. It is important to know that if you __sum__ two vectors in R, it takes the element-wise sum. For example, the following three statements are completely equivalent:

```{r eval = F}
c(1, 2, 3) + c(4, 5, 6)
c(1 + 4, 2 + 5, 3 + 6)
c(5, 7, 9)
```

### Types of Vectors

So there are there are four common types of vectors, depending on the class: 
* `logical` 
* `integer` 
* `numeric` (same as `double`)
* `character`.

#### Logical Vectors {-}

Logical vectors take on one of three possible values:

1. `TRUE`
2. `FALSE`
3. `NA` (missing value)

```{r}
c(TRUE, TRUE, FALSE, NA)
```

#### Numeric Vectors {-}

Numeric vectors contain numbers. They can be stored as *integers* (whole numbers) or *doubles* (numbers with decimal points). In practice, you rarely need to concern yourself with this difference, but just know that they are different but related things.

```{r}
c(1, 2, 335)
c(4.2, 4, 6, 53.2)
```

#### Character Vectors {-}

Character vectors contain character (or 'string') values. Note that each value has to be surrounded by quotation marks *before* the comma.

```{r}
c("Beyonce", "Kelly", "Michelle", "LeToya")
```

### Coercion

We can change or convert a vector's type using `as....`.

```{r}
num_var <- c(1, 2.5, 4.5)
class(num_var)
as.character(num_var)
```

Remember that all elements of a vector must be the same type. So when you attempt to combine different types, they will be __coerced__ to the most "flexible" type. 

For example, combining a character and an integer yields a character:

```{r}
c("a", 1)
```

Guess what the following do without running them first:

```{r eval = F}
c(1.7, "a") 
c(TRUE, 2) 
c("a", TRUE) 
```

#### TRUE == 1 and FALSE == 0. {-}

Notice that when a logical vector is coerced to an integer or double, `TRUE` becomes 1 and `FALSE` becomes 0. This is very useful in conjunction with `sum()` and `mean()`

```{r}
x <- c(FALSE, FALSE, TRUE)
as.numeric(x)

# Total number of TRUEs
sum(x)

# Proportion that are TRUE
mean(x)
```

#### Coercion often happens automatically. {-}

This is called _implicit coercion_. Most mathematical functions (`+`, `log`, `abs`, etc.) will coerce to a double or integer, and most logical operations (`&`, `|`, `any`, etc) will coerce to a logical. You will usually get a warning message if the coercion might lose information. 

```{r}
1 < "2"
"1" > 2
```

Sometimes coercions, especially nonsensical ones, won’t work.

```{r}
x <- c("a", "b", "c")
as.numeric(x)
as.logical(x)
```

### Challenges

#### Challenge 1: Create and examine your vector {-}

Create a character vector called `fruit` that contain 4 of your favorite fruits. Then evaluate its structure using the commands below.

```{r, eval = FALSE}

# First create your fruit vector 
# YOUR CODE HERE


# Examine your vector
length(fruit)
class(fruit)
str(fruit)
```

#### Challenge 2: Coercion {-}

```{r, eval=FALSE}

# 1. Create a vector of a sequence of numbers between 1 to 10.

# 2. Coerce that vector into a character vector

# 3. Add the element "11" to the end of the vector

# 4. Coerce it back to a numeric vector.
```

#### Challenge 3: Calculations on Vectors{-}

Create a vector of the numbers 11-20, and multiply it by the original vector from Challenge 2.

