---
title: "One-Dimensional Data in R"
author: "PLSC 31101"
date: "Fall 2019"
output: html_document
---

## Lists

Lists are different from vectors because their elements can be of __any type__. Lists are sometimes called recursive vectors, because a list can contain other lists. This makes them fundamentally different from vectors. 

### Creating Lists

You construct lists by using `list()` instead of `c()`: 

```{r}
x <- list(1, "a", TRUE, c(4, 5, 6))
x
```

### Naming Lists

As with vectors, we can attach names to each element on our list:

```{r eval = F}
my_list <- list(name1 = elem1, 
                name2 = elem2)
```

This creates a list with components that are named `name1`, `name2`, and so on. If you want to name your lists after you've created them, you can use the `names()` function as you did with vectors. The following commands are fully equivalent to the assignment above:

```{r eval = F}
my_list <- list(elem1, elem2)
names(my_list) <- c("name1", "name2")
```

### List Structure

A very useful tool for working with lists is `str()` because it focuses on reviewing the structure of a list, not the contents.

```{r}
x <- list(a = c(1, 2, 3),
          b = c("Hello", "there"),
          c = 1:10)
str(x)
```

A list does not print to the console like a vector. Instead, each element of the list starts on a new line.

```{r}
x.vec <- c(1,2,3)
x.list <- list(1,2,3)
x.vec
x.list
```

Lists are used to build up many of the more complicated data structures in R. For example, both data frames and linear models objects (as produced by `lm()`) are lists:

```{r}
head(mtcars)
is.list(mtcars)
mod <- lm(mpg ~ wt, data = mtcars)
is.list(mod)
```

You could say that a list is some kind super data type: you can store practically any piece of information in it!

For this reason, lists are extremely useful inside functions. You can "staple" together lots of different kinds of results into a single object that a function can return.

```{r}
mod <- lm(mpg ~ wt, data = mtcars)
str(mod)
```

### Challenges

#### Challenge 1. {-}

What are the four basic types of atomic vectors? How does a list differ from an atomic vector?

#### Challenge 2. {-}

Why is `1 == "1"` true? Why is `-1 < FALSE` true? Why is `"one" < 2` false?

#### Challenge 3. {-}

Create three vectors and combine them into a list. Assign them names.

#### Challenge 4.  {-}

If `x` is a list, what is the class of `x[1]`?  How about `x[[1]]`?
