---
title: "Two-Dimensional Data in R"
author: "PLSC 31101"
date: "Fall 2019"
output: html_document
---

## Matrices

Matrices are 2-d vectors. That is, they are a collection of elements of the same data type (numeric, character, or logical), arranged into a fixed number of rows and columns. 

By definition, if you want to combine different types of data (one column numbers, another column characters), you want a **dataframe**, not a matrix. 

### Creating Matrices

We can create a matrix using the `matrix()` function. In this function, we assign dimensions to a vector, like this:

```{r}
m <- matrix(1:6, nrow = 2, ncol = 3)
m
```

Notice that matrices fill column-wise. We can change this using the `byrow` argument:

```{r}
m <- matrix(1:6, byrow = T, nrow = 2, ncol = 3)
m
```

Another way to create matrices is to bind columns or rows using `cbind()` and `rbind()`.

```{r}
x <- 1:3
y <- 10:12
cbind(x, y)
# or 
rbind(x, y)
```

### Matrix Dimensions

Use `dim()` to find out how many rows or columns are in a matrix (or dataframe)

```{r}
dim(m)
```

We can transpose a matrix (or dataframe) with `t()`

```{r}
m <- matrix(1:6, nrow = 2, ncol = 3)
m
t(m)
```

### Matrix Names

Just like vectors or lists, we can give matrices names that describe the rows and columns

```{r}
m <- matrix(1:6, nrow = 2, ncol = 3)

rownames(m) <- c("row1", "row2")
colnames(m) <- c("A", "B", "C")

m
```

### Challenge

Take a look at the vector I've created about box office sales for the first three Harry Potter movies:

```{r}
# Box office sales (in millions!)
philosophers_stone <- c(66.1, 317.6, 657.2)
chamber_secrets <- c(54.7, 261.9, 616.9)
prisoner_azkaban <- c(45.6, 249.5, 547.1)

# Vectors region and titles, used for naming
region <- c("UK", "US", "Other")
titles <- c("Philosopher's Stone", "Chamber of Secrets", "Prisoner of Azkaban")
```

Your challenge is to:

1. Combine the first three vectors into a matrix
2. Add names for the matrix's rows (`titles`) and columns (`region`)
3. Use `rowSums()` to find the total Worldwide Box Office sales for each movie.


