---
title: "Classes"
author: "PLSC 31101"
date: "Fall 2019"
output: html_document
---

# Subsetting

When working with data, you'll need to subset objects early and often. Luckily, R's subsetting operators are powerful and fast. Mastery of subsetting allows you to succinctly express complex operations in a way that few other languages can match. Subsetting is hard to learn because you need to master a number of interrelated concepts:

* The three subsetting operators: `[`, `[[`, and `$`.

* The four types of subsetting.

* The important differences in behaviour for different objects (e.g., vectors, lists, factors, matrices, and data frames).

* The use of subsetting in conjunction with assignment.

This unit helps you master subsetting by starting with the simplest type of subsetting: subsetting an atomic vector with `[`. It then gradually extends your knowledge, first to more complicated data types (like dataframes and lists), and then to the other subsetting operators, `[[` and `$`. You'll then learn how subsetting and assignment can be combined to modify parts of an object, and, finally, you'll see a large number of useful applications.

## Subsetting Vectors

It's easiest to learn how subsetting works for vectors, and then how it generalises to higher dimensions and other more complicated objects. We'll start with `[`, the most commonly used operator.

### Subsetting Types

Let's explore the different types of subsetting with a simple vector, `x`. 

```{r}
x <- c(2.1, 4.2, 3.3, 5.4)
```

Note that the number after the decimal point gives the original position in the vector.

There are four things you can use to subset a vector: 

#### 1.   __Positive integers__ return elements at the specified positions: {-}

```{r}
(x <- c(2.1, 4.2, 3.3, 5.4))
x[1]
```

We can also index multiple values by passing a vector of integers:

```{r}
(x <- c(2.1, 4.2, 3.3, 5.4))
x[c(3, 1)]

# Duplicated indices yield duplicated values
x[c(1, 1)]
```

Note that you *have* to use `c` inside the `[` for this to work!

More examples:

```{r}
# `order(x)` gives the index positions of smallest to largest values.
(x <- c(2.1, 4.2, 3.3, 5.4))
order(x)

# use this to order values.
x[order(x)]
x[c(1, 3, 2, 4)]
```

#### 2.   __Negative integers__ omit elements at the specified positions: {-}

```{r}
x <- c(2.1, 4.2, 3.3, 5.4)
x[-1]
x[-c(3, 1)]
```

You can't mix positive and negative integers in a single subset:

```{r, error = TRUE}
x <- c(2.1, 4.2, 3.3, 5.4)
x[c(-1, 2)]
```

#### 3. __Character vectors__ return elements with matching names. This only works if the vector is named. {-}
 
```{r}
x <- c(2.1, 4.2, 3.3, 5.4)

# apply names
names(x) <- c("a", "b", "c", "d")

# subset using names
x[c("d", "c", "a")]

# Like integer indices, you can repeat indices
x[c("a", "a", "a")]

# Careful! Names are always matched exactly
x <- c(abc = 1, def = 2)
x[c("a", "d")]
```

#### 4.   __Logical vectors__ select elements where the corresponding logical value is `TRUE`. {-}

```{r}
x <- c(2.1, 4.2, 3.3, 5.4)
x[c(TRUE, TRUE, FALSE, FALSE)]
```

### Conditional Subsetting

Logical subsetting is the most useful type of subsetting, because you use it to subset based on __conditional__ or __comparative__ statements.

The (logical) comparison operators known to R are:

- `<` for less than
- `>` for greater than
- `<=` for less than or equal to
- `>=` for greater than or equal to
- `==` for equal to each other
- `!=` not equal to each other

The nice thing about R is that you can use these comparison operators also on vectors. For example:

```{r}
x <- c(2.1, 4.2, 3.3, 5.4)
x > 3
```

This command tests for every element of the vector if the condition stated by the comparison operator is `TRUE` or `FALSE`. And it returns a logical vector! 

We can now pass this statement between the square brackets that follow `x` to subset only those items that match `TRUE`:

```{r}
x[x > 3]
```

You can combine conditional statements with `&` (and), `|` (or), and `!` (not)

```{r}
x <- c(2.1, 4.2, 3.3, 5.4)

# combing two conditional statements with &
x > 3 & x < 5
x[x > 3 & x < 5]

# combing two conditional statements with |
x < 3 | x > 5 
x[x < 3 | x > 5]

# combining conditional statements with !
!x > 5 
x[!x > 5]
```

Another way to generate implicit conditional statements is using the `%in%` operator, which tests whether an item is in a set: 

```{r}
x <- c(2.1, 4.2, 3.3, 5.4)

# generate implicit logical vectors through the %in% operator
x %in% c(3.3, 4.2)
x[x %in% c(3.3, 4.2)]
```

### Challenge

Subset `country.vector` below to return every value EXCEPT "Canada" and "Brazil"

```{r}
country.vector<-c("Afghanistan", "Canada", "Sierra Leone", "Denmark", "Japan", "Brazil")

# Do it using positive integers

# Do it using negative integers

# Do it using a logical vector

# Do it using a conditional statement (and an implicit logical vector)
```