---
title: "Subsetting"
author: "PLSC 31101"
date: "Fall 2019"
output: html_document
---

```{r include = F}
library(kableExtra)
```

## Subsetting Lists 

Subsetting a list works in the same way as subsetting an atomic vector. However, there's one important difference: `[` will always return a list. `[[` and `$`, as described below, let you pull out the components of the list.

Let's illustrate with the following list `my_list`:

```{r}
my_list <- list(a = 1:3, b = "a string", c = pi, d = list(-1, -5))
```

### With `[`

`[` extracts a sub-list where the result will always be a list. Like with vectors, you can subset with a logical, integer, or character vector.


```{r}
my_list[1:2]
str(my_list[1:2])

my_list[4]
str(my_list[4])

my_list["a"]
str(my_list["a"])
```

### With `[[`

`[[` extracts a single _component_ from a list. In other words, it removes that hierarchy and returns whatever object is stored inside.

```{r}
my_list[[1]]
str(my_list[[1]])

# compare to
my_list[1]
str(my_list[1])
```
The distinction between `[` and `[[` is really important for lists, because `[[` drills down into the list while `[` returns a new, smaller list. 

>  "If list `x` is a train carrying objects, then `x[[5]]` is
> the object in car 5; `x[4:6]` is a train of cars 4-6." 
>
> --- @RLangTip


### with `$`

`$` is a shorthand for extracting named elements of a list. It works similarly to `[[` except that you don’t need to use quotes.

```{r}
my_list$a

# same as
my_list[["a"]]
```

The `$` operator becomes especially helpful when applied to dataframes, explained more below.

### Challenge

Take a look at the linear model below:

```{r}
mod <- lm(mpg ~ wt, data = mtcars)
summary(mod)
```

Extract the R squared from the model summary.

## Subsetting Matrices

Similar to vectors, you can use the square brackets `[ ]` to select one or multiple elements from a matrix. But whereas vectors have one dimension, matrices have two dimensions. We therefore have to use two subsetting vectors -- one for rows to select, another for columns -- separated by a comma.

Check out the following matrix:

```{r}
a <- matrix(1:9, nrow = 3)
colnames(a) <- c("A", "B", "C")
a
```

We can subset this matrix by passing two subsetting vectors: one to select rows, another to select columns: 

```{r}
# selects the value at the first row and second column
a[1, 2] 

# selects first row, and the first and third columns
a[1, -2] 

# selects first two rows, and the first and third columns
a[c(1,2), c(1, 3)] 
```

Blank subsetting is also useful because it lets you keep all rows or all columns.

```{r}
a[c(1, 2), ] # selects first two rows and all columns
```

