---
title: "Subsetting"
author: "PLSC 31101"
date: "Fall 2019"
output: html_document
---

```{r include = F}
library(kableExtra)
```

## Subsetting Dataframes

Data from data frames can be addressed like matrices, using two vectors separated by a comma.

```{r}
# Definition of vectors
name <- c("Mercury", "Venus", "Earth", "Mars", "Jupiter", "Saturn", "Uranus", "Neptune")
type <- c("Terrestrial planet", "Terrestrial planet", "Terrestrial planet", "Terrestrial planet", "Gas giant", "Gas giant", "Gas giant", "Gas giant")
diameter <- c(0.382, 0.949, 1, 0.532, 11.209, 9.449, 4.007, 3.883)
rings <- c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE)

planets <- data.frame(name, type, diameter, rings, stringsAsFactors = F)
planets
```

Let's try some subsetting now. 

```{r}
# Print out diameter of Mercury (row 1, column 3)
planets[1, 3]

# Print out data for Mars (entire fourth row)
planets[4, ]

# Print first two rows of the first two columns
planets[1:2, 1:2]
```

### Subsetting Names and `$`

Instead of using numerics to select elements of a data frame, you can also use the variable names to select columns of a data frame.

Suppose you want to select the first three elements of the type column. One way to do this is

```{r}
planets[1:3, 2]
```

A possible disadvantage of this approach is that you have to know (or look up) the column number of type, which gets hard if you have a lot of variables. It is often easier to just make use of the variable name:

```{r}
planets[1:3, "type"]
```

You will often want to select an entire column, namely one specific variable from a data frame. If you want to select all elements of the variable "diameter", for example, both of these will do the trick:

```{r}
planets[,3]
planets[,"diameter"]
```

However, there is a short-cut. If your columns have names, you can use the `$` sign:

```{r}
planets$diameter
```

Remember that datasets are really lists of vectors (one vector per column). Just as `list$name` selects the `name` element from the list, `df$name` selects the `name` column (vector) from the dataframe.

### Conditional Subsetting

What if we want to subset the dataset based on some condition? Let's say we want to extract all the planets with a diameter greater than 3? We could inspect the dataset and record all the observations that fit that description, but that's tedious and error prone.

There's a better way! We can combine two powerful subsetting tools: the `$` operator and conditional subsetting.

First, we extract the `diameter` column.

```{r}
diameters <- planets$diameter
```

Then, we find the elements that are greater than 3.

```{r}
diameters > 3
```

It's a boolean vector! We can now use this inside `[ , ]` to extract all plantes with `diameter > 3`. 

Think: Are we subsettings row or columns here?

```{r}
planets[diameters > 3, ]

# same as
# planets[planets$diameter > 3, ]
```

Because it allows you to easily combine conditions from multiple columns, logical subsetting is probably the most commonly used technique for extracting rows out of a data frame. 

### List-Like and Matrix-Like Subsetting

Data frames possess the characteristics of both lists and matrices: if you subset with a single vector, they behave like lists, and return only the columns.

```{r}
df <- data.frame(x = 4:6, y = 3:1, z = letters[1:3])

# Like a list:
df[c("x", "z")]

# Like a matrix
df[, c("x", "z")]
```

But there’s an important difference when you select a single column: matrix subsetting simplifies by default, list subsetting does not.

```{r}
df <- data.frame(x = 4:6, y = 3:1, z = letters[1:3])

# like a list
df["x"]
class(df["x"])

# like a matrix
df[, "x"]
class(df[, "x"])
```

### Challenges

#### Challenge 1. {-}

Fix each of the following common data frame subsetting errors:

```{r eval = F}
# check out what we're dealing with
mtcars

# fix
mtcars[mtcars$cyl = 4, ]
mtcars[-1:4, ]
mtcars[mtcars$cyl <= 5]
mtcars[mtcars$cyl == 4 | 6, ]
```

#### Challenge 2. {-} 

Why does `mtcars[1:20]` return an error? How does it differ from the similar `mtcars[1:20, ]`? 