---
title: "Classes"
author: "PLSC 31101"
date: "Fall 2019"
output: html_document
---

## Sub-assignment

### Basics of Sub-assignment

All subsetting operators can be combined with assignment to modify selected values of the input vector.

```{r, error = TRUE}
x <- 1:5
x[c(1, 2)] <- 2:3
x
```

This is especially useful when conditionally modifying vectors. For example, let's say we wanted to replace all values less than 3 with NA.

```{r}
x <- 1:5
x[x < 3] <- NA
x
```

This also works on dataframes. Let's say we wanted to modify our `planets` dataframe. 

```{r}
# Definition of vectors
name <- c("Mercury", "Venus", "Earth", "Mars", "Jupiter", "Saturn", "Uranus", "Neptune")
type <- c("Terrestrial planet", "Terrestrial planet", "Terrestrial planet", "Terrestrial planet", "Gas giant", "Gas giant", "Gas giant", "Gas giant")
diameter <- c(0.382, 0.949, 1, 0.532, 11.209, 9.449, 4.007, 3.883)
rings <- c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE)

planets <- data.frame(name, type, diameter, rings, stringsAsFactors = F)
planets
```

Let's say we want to replace the term "Terrestrial planet" with "TP". First we need to subset `type` for those elements:

```{r}
planets$type == "Terrestrial planet"
```

Now we can re-assign the values of `type`:

```{r}
planets$type[planets$type == "Terrestrial planet"]
planets$type[planets$type == "Terrestrial planet"] <- "TP"
planets
```


### Recycling

When applying an operation to two vectors that requires them to be the same length, R automatically recycles, or repeats, the shorter one, until it is long enough to match the longer one. 

```{r}
df <- data.frame(x = 4:7, y = letters[1:4])

# r recycles values
df$x <- c(1, 2)
df

# sometimes this is helpful if you want to replace an entire vector to one value.
df$x <- df$x + 3
df
```

### Applications

The basic principles described above give rise to a wide variety of useful applications. Some of the most important applications are described below. Many of these basic techniques are wrapped up into more concise functions (e.g., `subset()`, `merge()`, `plyr::arrange()`), but it is useful to understand how they are implemented with basic subsetting. This will allow you to adapt to new situations that are not dealt with by existing functions.

#### Ordering Columns {-}

Consider we have this data frame:

```{r}
df <- data.frame(
  Country = c("Iraq", "China", "Mexico", "Russia", "United Kingdom"),
  Region = c("Middle East", "Asia", "North America", "Eastern Europe", "Western Europe"),
  Language = c("Arabic", "Mandarin", "Spanish", "Russian", "English")
)
df
```

What if we wanted to reorder the columns so that `Region` is first? We can do so using subsetting with the names (or number) of the columns:

```{r}
df <- data.frame(
  Country = c("Iraq", "China", "Mexico", "Russia", "United Kingdom"),
  Region = c("Middle East", "Asia", "North America", "Eastern Europe", "Western Europe"),
  Language = c("Arabic", "Mandarin", "Spanish", "Russian", "English")
)

# reorder columns using names
names(df)
df1 <- df[, c("Region", "Country", "Language")]
df1

# reorder columns using indices
names(df)
df1 <- df[, c(2,1,3)]
df1
```

One helpul function is the `order` function. It takes a vector as input and returns an integer vector describing how the subsetted vector should be ordered:

```{r}
x <- c("b", "c", "a")
order(x)
x[order(x)]
```

Knowing this, we can use `order` to reorder our columns by alphabetical order.

#### Removing (or keeping) columns from data frames. {-}

There are two ways to remove columns from a data frame. You can set individual columns to `NULL`: 

```{r}
df <- data.frame(
  Country = c("Iraq", "China", "Mexico", "Russia", "United Kingdom"),
  Region = c("Middle East", "Asia", "North America", "Eastern Europe", "Western Europe"),
  Language = c("Arabic", "Mandarin", "Spanish", "Russian", "English")
)

df$Language <- NULL
```

Or you can subset to return only the columns you want:

```{r}
df <- data.frame(
  Country = c("Iraq", "China", "Mexico", "Russia", "United Kingdom"),
  Region = c("Middle East", "Asia", "North America", "Eastern Europe", "Western Europe"),
  Language = c("Arabic", "Mandarin", "Spanish", "Russian", "English")
)

df1 <- df[, c("Country", "Region")]
df1

# using negative integers
df2 <- df[, -3]
df2
```

