---
title: "Plotting and Visualization in R"
author: "PLSC 31101"
date: "Fall 2019"
output: html_document
---

## ggplot2

Setup:

```{r}
library(ggplot2)
gap <- read.csv("data/gapminder-FiveYearData.csv", stringsAsFactors = F)
```

#### Why `ggplot`? {-}

* More elegant & compact code than R base graphics 
* More aesthetically pleasing defaults than lattice
* Very powerful for exploratory data analysis
* Follows a grammar, just like any language.
* It defines basic components that make up a sentence. In this case, the grammar defines components in a plot.
* *G*rammar of *g*raphics originally coined by Lee Wilkinson

### Grammar

The general call for `ggplot2` looks like this:
  
```{r, eval=FALSE}
ggplot(data=, aes(x=, y=), color=, size=,) + geom_xxxx()+geom_yyyy()
```

The *grammar* involves some basic components:

1. **Data**: a data.frame
2. **Aes**thetics: How your data are represented visually, aka its "mapping". Which variables are shown on x, y axes, as well as color, size, shape, etc.
3. **Geom**etry: The geometric objects in a plot -- points, lines, polygons, etc.

The key to understanding `ggplot2` is thinking about a figure in layers: just like you might do in an image editing program like Photoshop, Illustrator, or Inkscape. 

Let's look at an example:

```
ggplot(data = gap, aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
```

So the first thing we do is call the `ggplot` function. This function lets R know that we’re creating a new plot, and any of the arguments we give the `ggplot` function are the global options for the plot: they apply to all layers on the plot.

We’ve passed in two arguments to `ggplot`. First, we tell `ggplot` what **`data`** we want to show on our figure, in this example the `gapminder` data we read in earlier. 

For the second argument we passed in the **`aes`** function, which tells `ggplot` how variables in the data map to aesthetic properties of the figure, in this case the x and y locations. Here we told `ggplot` we want to plot the `lifeExp` column of the gapminder data frame on the x-axis, and the `gdpPercap` column on the y-axis. 

Notice that we didn’t need to explicitly pass `aes` these columns (e.g. `x = gapminder[, "lifeExp""])`, this is because `ggplot` is smart enough to know to look in the data for that column!

By itself, the call to `ggplot` isn’t enough to draw a figure:

```{r eval=FALSE}
ggplot(data = gap, aes(x = gdpPercap, y = lifeExp))
```

We need to tell `ggplot` how we want to visually represent the data, which we do by adding a new **`geom`** layer. In our example, we used `geom_point`, which tells `ggplot` we want to visually represent the relationship between x and y as a scatterplot of points:

```{r}
ggplot(data = gap, aes(x = gdpPercap, y = lifeExp)) + 
  geom_point()

# same as
# my_plot <- ggplot(data = gap, aes(x = gdpPercap, y = lifeExp))
# my_plot + geom_point()
```

#### Challenge 1. {-}

Modify the example so that the figure visualises how life expectancy has changed over time:

Hint: the gapminder dataset has a column called `year`, which should appear on the x-axis.

### Anatomy of `aes`

In the previous examples and challenge we’ve used the `aes` function to tell the scatterplot `geom` about the **x** and **y** locations of each point. Another aesthetic property we can modify is the point **color**. 

```{r}
ggplot(data = gap, aes(x = gdpPercap, y = lifeExp, color=continent)) + 
  geom_point()
```

Normally, specifying options like `color="red"` or `size=10` for a given layer results in its contents being red and quite large.  Inside the `aes()` function, however, these arguments are given entire variables whose values will then be displayed using different realizations of that aesthetic.

**Color** isn't the only aesthetic argument we can set to display variation in the data. We can also vary by shape, size, etc.

```{r, eval=FALSE}
ggplot(data=, aes(x=, y=, by =, color=, linetype=, shape=, size=))
```

### Layers

In the previous challenge, you plotted `lifExp` over time. Using a scatterplot probably isn’t the best for visualising change over time. Instead, let’s tell `ggplot` to visualise the data as a line plot:

```{r}
ggplot(data = gap, aes(x=year, y=lifeExp, by=country, color=continent)) + 
  geom_line()
```

Instead of adding a `geom_point` layer, we’ve added a `geom_line` layer. We’ve also added the **by** aesthetic, which tells `ggplot` to draw a line for each country.

But what if we want to visualise both lines and points on the plot? We can simply add another layer to the plot:

```{r}
ggplot(data = gap, aes(x=year, y=lifeExp, by=country, color=continent)) + 
  geom_line() + 
  geom_point()
```

It’s important to note that each layer is drawn on top of the previous layer. In this example, the points have been drawn on top of the lines. Here’s a demonstration:

```{r}
ggplot(data = gap, aes(x=year, y=lifeExp, by=country)) + 
  geom_line(aes(color=continent)) + 
  geom_point()
```

In this example, the aesthetic mapping of **color** has been moved from the global plot options in `ggplot` to the `geom_line` layer so it no longer applies to the points. Now we can clearly see that the points are drawn on top of the lines.

#### Challenge 2. {-}

Switch the order of the point and line layers from the previous example. What happened?

### Labels 

Labels are considered to be their own layers in `ggplot`.

```{r}
# add x and y axis labels
ggplot(data = gap, aes(x = gdpPercap, y = lifeExp, color=continent)) + 
  geom_point() + 
  xlab("GDP per capita") + 
  ylab("Life Expectancy") + 
  ggtitle("My fancy graph")
```

So are scales:

```{r}
# limit x-axis from 1,000 to 20,000
ggplot(data = gap, aes(x = gdpPercap, y = lifeExp, color=continent)) + 
  geom_point() + 
  xlab("GDP per capita") + 
  ylab("Life Expectancy") + 
  ggtitle("My fancy graph") + 
  xlim(1000, 20000)
```

### Transformations and Stats

`ggplot` also makes it easy to overlay statistical models over the data. To demonstrate we’ll go back to an earlier example:

```{r}
ggplot(data = gap, aes(x = gdpPercap, y = lifeExp, color=continent)) + 
  geom_point()
```

We can change the scale of units on the x-axis using the `scale` functions. These control the mapping between the data values and visual values of an aesthetic.

```{r}
ggplot(data = gap, aes(x = gdpPercap, y = lifeExp, color=continent)) + 
  geom_point() + 
  scale_x_log10()
```

The `log10` function applied a transformation to the values of the `gdpPercap` column before rendering them on the plot, so that each multiple of 10 now only corresponds to an increase in 1 on the transformed scale, e.g. a GDP per capita of 1,000 is now 3 on the y axis, a value of 10,000 corresponds to 4 on the x-axis and so on. This makes it easier to visualise the spread of data on the x-axis.

We can fit a simple relationship to the data by adding another layer, `geom_smooth`:

```{r}
ggplot(data = gap, aes(x = gdpPercap, y = lifeExp, color=continent)) + 
  geom_point() + 
  scale_x_log10() + 
  geom_smooth(method="lm")
```

Note that we have 5 lines, one for each region, because of the **color** option is the global `aes` function. But if we move it, we get different results:

```{r}
ggplot(data = gap, aes(x = gdpPercap, y = lifeExp)) + 
  geom_point(aes(color=continent)) + 
  scale_x_log10() + 
  geom_smooth(method="lm")
```

So, there are two ways an aesthetic can be specified. Here, we set the **color** aesthetic by passing it as an argument to `geom_point`. Previously in the lesson, we used the `aes` function to define a *mapping* between data variables and their visual representation.

We can make the line thicker by setting the **size** aesthetic in the geom_smooth layer:

```{r}
ggplot(data = gap, aes(x = gdpPercap, y = lifeExp)) + 
  geom_point(aes(color=continent)) + 
  scale_x_log10() + 
  geom_smooth(method="lm", size = 1.5)
```

#### Challenge 3. {-}

Modify the color and size of the points on the point layer in the previous example so that they are fixed (i.e. not reflective of continent).

Hint: do not use the `aes` function.

### Facets

Earlier, we visualised the change in life expectancy over time across all countries in one plot. Alternatively, we can split this out over multiple panels by adding a layer of **facet** panels:

```{r}
ggplot(data = gap, aes(x = year, y = lifeExp)) +
  geom_point() + 
  facet_wrap( ~ continent)
```

### Putting Everything Together

Here are some other common `geom` layers:

**bar plots**

```{r}
# count of lifeExp bins
ggplot(data = gap, aes(x = lifeExp)) + 
  geom_bar(stat="bin")

# with color representing regions
ggplot(data = gap, aes(x = lifeExp, fill = continent)) + 
  geom_bar(stat="bin")
```

**box plots**

```{r}
ggplot(data = gap, aes(x = continent, y = lifeExp)) + 
  geom_boxplot()
```

This is just a taste of what you can do with `ggplot2`. 

RStudio provides a really useful [cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) of the different layers available, and more extensive documentation is available on the [ggplot2 website](http://docs.ggplot2.org/current/). 

Finally, if you have no idea how to change something, a quick Google search will usually send you to a relevant question and answer on Stack Overflow with reusable code to modify!

#### Challenge 4. {-}

Create a density plot of GDP per capita, filled by continent.

Advanced: 
- Transform the x-axis to better visualise the data spread. 
- Add a facet layer to panel the density plots by year.

## Saving plots

There are two basic image types:

1) **Raster/Bitmap** (.png, .jpeg)

Every pixel of a plot contains its own separate coding; not so great if you want to resize the image.

```{r, eval=FALSE}
jpeg(filename="example.png", width=, height=)
plot(x,y)
dev.off()
```

2) **Vector** (.pdf, .ps)

Every element of a plot is encoded with a function that gives its coding conditional on several factors; this is great for resizing.

```{r, eval=FALSE}
pdf(filename="example.pdf", width=, height=)
plot(x,y)
dev.off()
```

**Exporting with ggplot**

```{r, eval=FALSE}
# Assume we saved our plot is an object called example.plot
ggsave(filename="example.pdf", plot=example.plot, scale=, width=, height=)
```
