---
title: "Dictionary Methods"
author: "PLSC 31101"
date: "Fall 2019"
output: html_document
---

## Sentiment Analysis and Dictionary Methods

To demonstrate sentiment analysis, we're going to explore lyrics from Taylor Swift songs. 

Road the code below to get started.

```{r message = F}
require(tm)
require(tidytext)
require(tidyverse)
require(stringr)
require(textdata)

ts <- read.csv("data/taylor_swift.csv")
getwd()
```

### Preprocessing and Setup

First we must preprocess the corpus. Create a document-term matrix from the `lyrics` column of the `ts` data frame. Complete the following preprocessing steps:
- convert to lower
- remove stop words
- remove numbers
- remove punctuation.

**Think**: Why is stemming inappropriate for this application?

```{r}
# preprocess and create DTM
docs <- Corpus(VectorSource(ts$lyrics))

dtm <- DocumentTermMatrix(docs,
           control = list(tolower = TRUE,
                          removeNumbers = TRUE,
                          removePunctuation = TRUE,
                          stopwords = TRUE
                         ))

dtm <- as.data.frame(as.matrix(dtm))
```

#### Sentiment dictionaries {-}

We're going to use sentiment dictionaries from the `tidytext` package. Using the `get_sentiments` function, load the "bing" dictionary and store it in an object called `sent`. 

```{r}
sent <- get_sentiments("bing")
head(sent)
```

We'll now add a column to `sent` called `score`. This column should hold a "1" for positive words and "-1" for negative words.

```{r}
sent$score <- ifelse(sent$sentiment=="positive", 1, -1)
```

### Scoring the songs

We're now ready to score each song. 

(**NB**: There are probably many ways to program a script that performs this task. If you can think of a more elegant way, go for it!)

First, we'll create a dataframe that holds all the words in our dtm along with their sentiment score.

```{r}
# get all the words in our dtm and put it in a dataframe
words = data.frame(word = colnames(dtm))
head(words)

# get their sentiment scores
words <- merge(words, sent, all.x = T)
head(words)

# replace NAs with 0s
words$score[is.na(words$score)] <- 0
head(words)
```

We can now use matrix algebra (!!) to multiply our dtm by the scoring vector. This will return to us a score for each document (i.e., song).

```{r}
# calculate documents scores with matrix algebra! 
scores <- as.matrix(dtm) %*% words$score

# put it in the original documents data frame
ts$sentiment <- scores
```

Which song is happiest? Go listen to the song and see if you agree.

### Challenges

#### Challenge 1 {-} 

Using the code we wrote above, make a function that accepts 1) a vector of texts, and 2) a sentiment dictionary (i.e. a data frame with words and scores), and returns a vector of sentiment scores for each text.

```{r eval = F}
sentiment_score <- function(texts, sent_dict){
  
  # YOUR CODE HERE
  
  return(scores)
}
  
# uncomment to test it out!
# sentiment_score(ts$lyrics, sent)
```

#### Challenge 2 {-}

Using the function you wrote above, find out what the most and least positive Taylor Swift album is.

```{r eval = F}
# concatenate to make albums
albums <- ts %>% 
  group_by(album) %>%
  summarise(lyrics = str_c(lyrics, collapse = ";"))

# first load the dictionary
afinn <- get_sentiments("afinn")

# then run the function
sentiment_score(albums$lyrics, afinn)

# add to original df
albums$sent <- sentiment_score(albums$lyrics, afinn)
```

